---
title: "PracticalMachineLearning"
output: html_document
---
## Practical Machine Learning - Prediction Assignment Writeup

### Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large  
amount of data about personal activity relatively inexpensively. These type of devices are part  
of the quantified self movement – a group of enthusiasts who take measurements about themselves  
regularly to improve their health, to find patterns in their behavior, or because they are tech   
geeks. One thing that people regularly do is quantify how much of a particular activity they do,   
but they rarely quantify how well they do it. In this project, your goal will be to use data from   
accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to  
perform barbell lifts correctly and incorrectly in 5 different ways. More information is available  
from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting   
Exercise Dataset).  

### Data
The training data for this project is downloaded from the below -  
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv  

The test data is downloaded from the below:   
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv  

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har.   

### Description
Predictor variable -- classe.  
We would use seed value of 1234 to make this reproducible.  

#### Load the training data -
```{r, echo=TRUE}
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", destfile = "pml-training.csv", method="curl")
pml_training <- read.csv("pml-training.csv", header = TRUE)
```

#### Check summary statistics 
``` { r, echo=TRUE }
summary(pml_training)
str(pml_training, list.len = 999)
dim(pml_training)
(colSums(is.na(pml_training))) / (dim(pml_training)[1])
```

#### Observations of data
Two main points are observed in the data set -  
A) There are a lot of columns with NAs in it. So, we would find those columns and filter them.  
B) Some numeric data have been imported as factor because of the presence of some characters ("#DIV/0!")  

##### Data cleaning for "#DIV/0!" 
We would reload the data and load #DIV/0! as NA  

```{r, echo=TRUE }
pml_training <- read.csv("pml-training.csv", header = TRUE, na.strings=c("#DIV/0!"))
```

##### Data cleaning for missing data  
```{r , echo=TRUE}
checkforna <- (colSums(is.na(pml_training))) / (dim(pml_training)[1])
names(which(checkforna > 0.5 ))
pml_training[,names(which(checkforna > 0.5 ))] <- list(NULL)
```

##### Data cleaning for some variables 
Variables like "user_name" , "X",  "raw_timestamp_part_1" , "raw_timestamp_part_2" and cvtd_timestamp seem  
to be of no use and can be dropped.  

```{r , echo=TRUE}
pml_training[ , c("user_name", "X", "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp" )] <- list(NULL)
```

##### Load relevant libraries
```{r , echo=TRUE}
library(randomForest)
library(caret)
```

##### Convert factor variables to numerics ( except the predicted variable)
```{r, echo=TRUE}
classe <- pml_training$classe
pml_training$classe <- NULL
for (i in colnames(pml_training)) { pml_training[,i] <- as.numeric(pml_training[,i]) }
pml_training$classe <- classe
```

##### Plot classe variable for analysis
```{r, echo=TRUE}
plot(pml_training$classe)
```
No skew detected, so we can directly create partitions (no need to do stratified sampling)

#### Create data partitions and training/testing data sets
Setting a seed value to 1234 to make things reproducible.
70% data is for training and 30% for testing.

```{r , echo=TRUE}
set.seed(1234)
inTrain <- createDataPartition(pml_training$classe, p=0.7, list=FALSE)
training <- pml_training[inTrain,]
testing <- pml_training[-inTrain,]
```

#### Build a model using Random Forest
Tried different values for the number of trees and found that ntree of 200 gives fairly accurate   
results.  
Remember that more the number of trees, more is the computation involved in model building and  
predictions.  
```{r , echo=TRUE}
RF_model <- randomForest( classe ~ ., data=training, importance=TRUE, ntree=200)
```

##### Check the model for in-sample data
```{ r, echo=TRUE}
RF_model
```
Confusion matrix indicates OOB estimates ( for in-sample error ) of 0.33% which looks fairly accurate.  
The confusion matrix also indicate that predictions are accurate, so we move ahead with this model.  

##### Check the model for testing data
```{ r, echo=TRUE}
predictions <- predict( RF_model, newdata=testing[,-(dim(testing)[2]) ])
confusionMatrix(predictions, testing$classe )
```

ConfusionMatrix indicates accuracy of 0.9978 which is very accurate.  
The above details show very good values of Sensitivity and Specificity and hence indicate that  
the model is good.  
Predictions also look very accurate

#### Run predictions on new data set using the model built
We do the same data cleaning steps as we did in training and building a model.  
```{r, echo=TRUE}
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", destfile = "pml-testing.csv", method="curl")
pml_testing <- read.csv("pml-testing.csv", header = TRUE, na.strings=c("#DIV/0!"))
pml_testing[,names(which(checkforna > 0.5 ))] <- list(NULL)
pml_testing[ ,c("user_name", "X", "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp" )] <- list(NULL)
for (i in colnames(pml_testing)) { pml_testing[,i] <- as.numeric(pml_testing[,i]) }

predictions_new <- predict(RF_model, newdata = pml_testing[,-(dim(pml_testing)[2])])
```

##### Result of predictions on the new data set
```{r, echo=TRUE}
predictions_new
```

##### Write the Prediction to files (inside a folder called "results")
```{r, echo=TRUE}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_", i ,".txt")
    write.table(x[i], file = filename, quote = FALSE,
                row.names = FALSE, col.names = FALSE)
  }
}

setwd("results/");
pml_write_files(predictions_new)
```
